The Leaky Semicolon (Supplementary Material)

### 2021-12-30 paper/intro.tex
Fixing typos

For optimized concurrent languages, it is surprising difficult to define a
model that allows common compiler optimizations and hardware reorderings yet
disallows nonsense behaviors that don't arise in practice.  The latter are
commonly known as ``thin-air'' behaviors \cite{DBLP:conf/esop/BattyMNPS15}.
There are only a handful of solutions, and all have deficiencies.  These can
be classified by their approach to dependency tracking (from strongest to
weakest):
\begin{itemize}
\item Syntactic dependencies
  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}.
  These models require inefficient implementation of relaxed access.  This is
  a non-starter for safe languages like Java and Javascript, and may be an
  unacceptable cost for low-level languages like \cXI{}.
\item Semantic dependencies \cite{%
    Manson:2005:JMM:1047659.1040336, DBLP:conf/esop/JagadeesanPR10,
    DBLP:conf/popl/KangHLVD17, DBLP:journals/pacmpl/ChakrabortyV19,
    DBLP:conf/pldi/LeeCPCHLV20, promising-ldrf}. These models compute
  dependencies operationally using alternate worlds, making it impossible to
  understood a single execution in isolation; they also allow executions that
  violate temporal reasoning (see \textsection\ref{sec:related}).
\item No dependencies, as in \cXI{} \cite{DBLP:conf/esop/BattyMNPS15} and
  Javascript \cite{DBLP:journals/pacmpl/WattRP19}.  This allows thin-air
  executions.
\end{itemize}

@@

Part of this non-compositionality is essential: In a concurrent system, the
complete set of writes is known only at top-level.  However, much of it is
incidental.  Two recent models have attempted to limit non-compositionality.
\citet{DBLP:journals/pacmpl/JagadeesanJR20} defined Pomsets with
Preconditions (\PwP), which use preconditions and logic to calculate
dependencies for a Java-like language.
\citet{DBLP:conf/esop/PaviottiCPWOB20} defined Modular Relaxed Dependencies
(\MRD), which use event structures to calculate a semantic dependency
relation ($\rsdep$).  \PwP{} is defined using (acyclic) labelled partial
orders, or \emph{pomsets} \cite{GISCHER1988199}.  \MRD{} adds a causality
axiom to \cXI{}, stating that $({\rsdep}\cup{\rrfx})$ must be acyclic.  In
both approaches, acyclicity enables inductive reasoning.

@@

% % We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
% % be added to pomsets with
% % preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
% % compositional semantics for sequential composition.
% \begin{itemize}
% \item \S\ref{sec:model} presents the basic model, which satisfies many
%   desiderata, but not all.
% \item \S\ref{sec:arm} shows two approaches for efficient implementation on
%   Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
%   uses an optimal lowering, but requires a nontrivial change to the
%   definition of sequential composition.
% \item \S\ref{sec:additional} generalizes the basic semantics of read and write
%   to validate compiler optimizations.
% \end{itemize}
% % Acknowledgements go here, once we're not double-blinded.
% % The definitions in this paper have been formalized in Agda.
% Because it is closely related, we expect that the memory-model results of
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
% compositional reasoning for temporal safety properties and {local} \drfsc{}
% as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
% % In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
% % efficient compilation to \armeight{}, which improves upon that of
% % \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
% % characterization of \armeight{}.

@@

% \citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
% semicolon'' is problematic for programmers, and thus concurrent languages
% should guarantee sequential abstraction.  But processor and language
% implementors balk at the prospect, due to significant costs.

@@

This model extends naturally to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example
(where $\SEMI\!$ has higher precedence than $\PAR$):
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \wki{wx0}{wx1}
      \wki{wx1}{wy2}
      \wki{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}
We can represent such an execution as a labelled partial order, or
\emph{pomset} \cite{DBLP:conf/lop/Pratt85,GISCHER1988199}.  A program may give rise to many
executions, each reflecting a different interleaving of the threads.

@@

Our first attempt is to associate a
predicate transformer with each pomset. We visualize this in diagrams by
showing how $\bForm$ is transformed, for example:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{1}}{}
        \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of rx0}
        %\xo{rx0}{rx0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \PR{y}{s}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{1}}{}
        \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of ry0}
        %\xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{wz0}{(s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
        \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
        %\xo{wz0}{wz0d}
      \end{tikzinline}}
  \end{gathered}
\end{align*}
The predicate transformer for a write $\PW{z}{\aExp}$ matches
\citeauthor{DBLP:journals/cacm/Dijkstra75}: taking $\bForm$ to
$\bForm[\aExp/z]$.  For a read $\PR{\aLoc}{\aReg}$, however,
\citeauthor{DBLP:journals/cacm/Dijkstra75} would transform $\bForm$ to
$\bForm[x/r]$, which is equivalent to $(x{=}r) \limplies \bForm$ under the
assumption that registers are assigned at most once.  Instead, we use
$(1{=}r) \limplies \bForm$, reflecting the fact that $1$ may come from a
concurrent write.  The obligation to find a matching write is moved from the
sequential semantics of \emph{substitution} and \emph{implication} to the
concurrent semantics of \emph{fulfillment}.

@@

For a sequentially consistent semantics, sequential composition is
straightforward: we apply each predicate transformer to subsequent
preconditions, composing the predicate transformers.
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \tag{*}\label{eq1}
  \\
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{rx0}{\DR{x}{1}}{}
      \event{ry0}{\DR{y}{1}}{right=of rx0}
      \event{wz0}{(1{=}r) \limplies (1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm[r{*}(s{-}1)/z]}{right=of wz0}
      \wki{rx0}{ry0}
      \wki{ry0}{wz0}
      %\xo{rx0}{rx0ry0d}
      %\xo[out=155,in=15]{ry0}{rx0ry0d}
      % \xo{wz0}{rx0ry0d}
  \end{tikzinline}}
\end{gather*}
This works for the sequentially consistent case, but needs to be
weakened for the relaxed case.

The key observation of this paper is
that rather than working with one predicate transformer, we should
work with a \emph{family} of predicate transformers, indexed by sets
of events.
For example, for single-event pomsets, there are two predicate
transformers, since there are two subsets of any one-element set.
The \emph{independent}
transformer is indexed by the empty set, whereas the \emph{dependent}
transformer is indexed by the singleton.
We visualize this by including more than one transformed predicate,
with a dotted edge leading to the dependent one ($\xxo$). For example:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{1}}{}
        \xform{rx0i}{\bForm}{left=.4em of rx0}
        \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of rx0}
        \xo{rx0}{rx0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \PR{y}{s}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{1}}{}
        \xform{ry0i}{\bForm}{left=.4em of ry0}
        \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of ry0}
        \xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gathered}
  % &&
  % \begin{gathered}
  %   \IF{s{<}1} \THEN \PW{z}{r{*}s} \FI
  %   \\
  %   \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
  %     \event{wz0}{(s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
  %     \xform{wz0i}{\bForm}{left=.4em of wz0}
  %     \xform{wz0d}{\bForm}{right=of wz0}
  %     \xo{wz0}{wz0d}
  %   \end{tikzinline}}
  % \end{gathered}
\end{align*}
The model of sequential composition then picks which
predicate transformer to apply to an event's precondition by picking
the one indexed by all the events before it in causal order.

For example, we can recover the expected semantics
for \eqref{eq1} by choosing
the predicate transformer which is independent of $(\DR x1)$
but dependent on $(\DR y1)$, which is the transformer
which maps $\bForm$ to $(1{=}s) \limplies \bForm$.
(In subsequent diagrams, we only show predicate transformers for reads.)
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=2ex and 1.2em]
      \xform{wz0i}{\bForm}{}
      \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of wz0i}
      \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm}{right=of rx0d}
      \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of rx0ry0d}
      \event{rx0}{\DR{x}{1}}{above left=of rx0ry0d}
      \event{ry0}{\DR{y}{1}}{above right=of rx0ry0d}
      \event{wz0}{(1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
      \xo{rx0}{rx0d}
      \xo{ry0}{ry0d}
      \xo{rx0}{rx0ry0d}
      \xo{ry0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
In the diagram, the dotted lines indicate set inclusion into the index of the transformer-family.
As a quick correctness check, we can see that sequential composition is
associative in this case, since it does not matter whether we
associate to the left---with the intermediate step as in the diagram above,
eliding the write action---or to the right---with the intermediate step:
% \begin{gather*}
%   \PR{x}{r}\SEMI \PR{y}{s}
%   \\
%   \hbox{\begin{tikzinline}[node distance=2ex and 1.5em]
%       \xform{rx0d}{(0{=}r) \limplies \bForm}{}
%       \event{rx0}{\DR{x}{0}}{right=of rx0d}
%       \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
%       \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
%       \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
%       \xform{wz0i}{\bForm}{left=of rx0d}
%       \xo{rx0}{rx0d}
%       \xo{ry0}{ry0d}
%       \xo{rx0}{rx0ry0d}
%       \xo{ry0}{rx0ry0d}
%     \end{tikzinline}}
% \end{gather*}
\begin{gather*}
  \PR{y}{s}\SEMI \IF{s{<}1} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{ry0}{\DR{y}{1}}{}
      \event{wz0}{(1{=}s) \limplies (s {<} 1) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \xform{ry0d}{(1{=}s) \limplies \bForm}{left=of ry0}
      \xform{wz0i}{\bForm}{left=of ry0d}
      \po{ry0}{wz0}
      \xo{ry0}{ry0d}
    \end{tikzinline}}
\end{gather*}
This is an instance of the general result that sequential composition forms a
monoid (\reflem{lem:monoid}).

>>>>

For optimized concurrent languages, it is surprising difficult to define a
model that allows common compiler optimizations and hardware reorderings yet
disallows nonsense behaviors that don't arise in practice.  The latter are
commonly known as ``thin-air'' behaviors \cite{DBLP:conf/esop/BattyMNPS15}.
There are only a handful of solutions, and all have deficiencies.  These can
be classified by their approach to dependency tracking (from strongest to
weakest):
\begin{itemize}
\item Syntactic dependencies
  \cite{DBLP:conf/pldi/LahavVKHD17,Boehm:2014:OGA:2618128.2618134,DBLP:conf/oopsla/VafeiadisN13,DBLP:journals/corr/abs-1804-04214}.
  These models require inefficient implementation of relaxed access.  This is
  a non-starter for safe languages like Java and JavaScript, and may be an
  unacceptable cost for low-level languages like \cXI{}.
\item Semantic dependencies \cite{Manson:2005:JMM:1047659.1040336,DBLP:conf/esop/JagadeesanPR10,DBLP:conf/popl/KangHLVD17,DBLP:journals/pacmpl/ChakrabortyV19,DBLP:conf/pldi/LeeCPCHLV20,promising-ldrf}. These models compute
  dependencies operationally using alternate worlds, making it impossible to
  understood a single execution in isolation; they also allow executions that
  violate temporal reasoning (see \textsection\ref{sec:related}).
\item No dependencies, as in \cXI{} \cite{DBLP:conf/esop/BattyMNPS15} and
  JavaScript \cite{DBLP:journals/pacmpl/WattRP19}.  This allows thin-air
  executions.
\end{itemize}

@@

Part of this non-compositionality is essential: In a concurrent system, the
complete set of writes is known only at top-level.  However, much of it is
incidental.  Two recent models have attempted to limit non-compositionality.
\citet{DBLP:journals/pacmpl/JagadeesanJR20} defined Pomsets with
Preconditions (\PwP), which use preconditions and logic to calculate
dependencies for a Java-like language.
\citet{DBLP:conf/esop/PaviottiCPWOB20} defined Modular Relaxed Dependencies
(\MRD), which use event structures to calculate a semantic dependency
relation ($\rsdep$).  \PwP{} is defined using (acyclic) labeled partial
orders, or \emph{pomsets} \cite{GISCHER1988199}.  \MRD{} adds a causality
axiom to \cXI{}, stating that $({\rsdep}\cup{\rrfx})$ must be acyclic.  In
both approaches, acyclicity enables inductive reasoning.

@@

% % We show how predicate transformers~\cite{DBLP:journals/cacm/Dijkstra75} can
% % be added to pomsets with
% % preconditions~\cite{DBLP:journals/pacmpl/JagadeesanJR20} to create a
% % compositional semantics for sequential composition.
% \begin{itemize}
% \item \S\ref{sec:model} presents the basic model, which satisfies many
%   desiderata, but not all.
% \item \S\ref{sec:arm} shows two approaches for efficient implementation on
%   Arm.  The first uses a suboptimal lowering for acquiring reads.  The second
%   uses an optimal lowering, but requires a nontrivial change to the
%   definition of sequential composition.
% \item \S\ref{sec:additional} generalizes the basic semantics of read and write
%   to validate compiler optimizations.
% \end{itemize}
% % Acknowledgments go here, once we're not double-blinded.
% % The definitions in this paper have been formalized in Agda.
% Because it is closely related, we expect that the memory-model results of
% \cite{DBLP:journals/pacmpl/JagadeesanJR20} apply to our model, including
% compositional reasoning for temporal safety properties and {local} \drfsc{}
% as in \cite{Dolan:2018:BDR:3192366.3192421,DBLP:conf/ppopp/DongolJR19,promising-ldrf}.
% % In \textsection\ref{sec:arm}, we provide an alternative proof strategy for
% % efficient compilation to \armeight{}, which improves upon that of
% % \cite{DBLP:journals/pacmpl/JagadeesanJR20} by using a recent alternative
% % characterization of \armeight{}.

@@

% \citet{DBLP:conf/snapl/MarinoMMNS15} argue that the ``silently shifting
% semicolon'' is problematic for programmers, and thus concurrent languages
% should guarantee sequential abstraction.  But processor and language
% implementers balk at the prospect, due to significant costs.

@@

This model extends naturally to the case of shared-memory concurrency, leading to a \emph{sequentially consistent}
semantics \cite{Lamport:1979:MMC:1311099.1311750}, in which \emph{program order} inside a thread implies
a total \emph{causal order} between read and write events, for example
(where $\SEMI\!$ has higher precedence than $\PAR$):
\begin{gather*}
  \THREAD{\PW{x}{0}\SEMI \PW{x}{1}\SEMI \PW{y}{2}}
  \PAR
  \THREAD{\PR{y}{r}\SEMI \PR{x}{s}}
  \\[-.4ex]
  \hbox{\begin{tikzinline}[node distance=1.5em]
      \event{wx0}{\DW{x}{0}}{}
      \event{wx1}{\DW{x}{1}}{right=of wx0}
      \event{wy2}{\DW{y}{2}}{right=of wx1}
      \event{ry2}{\DR{y}{2}}{right=3em of wy2}
      \event{rx1}{\DR{x}{1}}{right=of ry2}
      \rf[out=15,in=165]{wy2}{ry2}
      \rf[out=15,in=165]{wx1}{rx1}
      \wki{wx0}{wx1}
      \wki{wx1}{wy2}
      \wki{ry2}{rx1}
    \end{tikzinline}}
\end{gather*}
We can represent such an execution as a labeled partial order, or
\emph{pomset} \cite{DBLP:conf/lop/Pratt85,GISCHER1988199}.  A program may give rise to many
executions, each reflecting a different interleaving of the threads.

@@

Our first attempt is to associate a
predicate transformer with each pomset. We visualize this in diagrams by
showing how $\bForm$ is transformed, for example:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{1}}{}
        \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of rx0}
        %\xo{rx0}{rx0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \PR{y}{s}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{1}}{}
        \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of ry0}
        %\xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \IF{s} \THEN \PW{z}{r{*}(s{-}1)} \FI
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{wz0}{(s{\neq}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
        \xform{wz0d}{\bForm[r{*}s/z]}{right=of wz0}
        %\xo{wz0}{wz0d}
      \end{tikzinline}}
  \end{gathered}
\end{align*}
The predicate transformer for a write $\PW{z}{\aExp}$ matches
\citeauthor{DBLP:journals/cacm/Dijkstra75}: taking $\bForm$ to
$\bForm[\aExp/z]$.  For a read $\PR{\aLoc}{\aReg}$, however,
\citeauthor{DBLP:journals/cacm/Dijkstra75} would transform $\bForm$ to
$\bForm[x/r]$, which is equivalent to $(x{=}r) \limplies \bForm$ under the
assumption that registers are assigned at most once.  Instead, we use
$(1{=}r) \limplies \bForm$, reflecting the fact that $1$ may come from a
concurrent write.  The obligation to find a matching write is moved from the
sequential semantics of \emph{substitution} and \emph{implication} to the
concurrent semantics of \emph{fulfillment}.

@@

For a sequentially consistent semantics, sequential composition is
straightforward: we apply each predicate transformer to subsequent
preconditions, composing the predicate transformers.
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \tag{*}\label{eq1}
  \\
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{rx0}{\DR{x}{1}}{}
      \event{ry0}{\DR{y}{1}}{right=of rx0}
      \event{wz0}{(1{=}r) \limplies (1{=}s) \limplies (s{\neq}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm[r{*}(s{-}1)/z]}{right=of wz0}
      \wki{rx0}{ry0}
      \wki{ry0}{wz0}
      %\xo{rx0}{rx0ry0d}
      %\xo[out=155,in=15]{ry0}{rx0ry0d}
      % \xo{wz0}{rx0ry0d}
  \end{tikzinline}}
\end{gather*}
This works for the sequentially consistent case, but needs to be
weakened for the relaxed case.

The key observation of this paper is
that rather than working with one predicate transformer, we should
work with a \emph{family} of predicate transformers, indexed by sets
of events.
For example, for single-event pomsets, there are two predicate
transformers, since there are two subsets of any one-element set.
The \emph{independent}
transformer is indexed by the empty set, whereas the \emph{dependent}
transformer is indexed by the singleton.
We visualize this by including more than one transformed predicate,
with a dotted edge leading to the dependent one ($\xxo$). For example:
\begin{align*}
  \begin{gathered}
    \PR{x}{r}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{rx0}{\DR{x}{1}}{}
        \xform{rx0i}{\bForm}{left=.4em of rx0}
        \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of rx0}
        \xo{rx0}{rx0d}
      \end{tikzinline}}
  \end{gathered}
  &&
  \begin{gathered}
    \PR{y}{s}
    \\
    \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
        \event{ry0}{\DR{y}{1}}{}
        \xform{ry0i}{\bForm}{left=.4em of ry0}
        \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of ry0}
        \xo{ry0}{ry0d}
      \end{tikzinline}}
  \end{gathered}
  % &&
  % \begin{gathered}
  %   \IF{s} \THEN \PW{z}{r{*}s} \FI
  %   \\
  %   \hbox{\begin{tikzinline}[node distance=1ex and1.5em]
  %     \event{wz0}{(s{\neq}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{}
  %     \xform{wz0i}{\bForm}{left=.4em of wz0}
  %     \xform{wz0d}{\bForm}{right=of wz0}
  %     \xo{wz0}{wz0d}
  %   \end{tikzinline}}
  % \end{gathered}
\end{align*}
The model of sequential composition then picks which
predicate transformer to apply to an event's precondition by picking
the one indexed by all the events before it in causal order.

For example, we can recover the expected semantics
for \eqref{eq1} by choosing
the predicate transformer which is independent of $(\DR x1)$
but dependent on $(\DR y1)$, which is the transformer
which maps $\bForm$ to $(1{=}s) \limplies \bForm$.
(In subsequent diagrams, we only show predicate transformers for reads.)
\begin{gather*}
  \PR{x}{r}\SEMI \PR{y}{s}\SEMI \IF{s} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=2ex and 1.2em]
      \xform{wz0i}{\bForm}{}
      \xform{rx0d}{(1{=}r) \limplies \bForm}{right=of wz0i}
      \xform{rx0ry0d}{(1{=}r) \limplies(1{=}s) \limplies \bForm}{right=of rx0d}
      \xform{ry0d}{(1{=}s) \limplies \bForm}{right=of rx0ry0d}
      \event{rx0}{\DR{x}{1}}{above left=of rx0ry0d}
      \event{ry0}{\DR{y}{1}}{above right=of rx0ry0d}
      \event{wz0}{(1{=}s) \limplies (s{\neq}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \po{ry0}{wz0}
      \xo{rx0}{rx0d}
      \xo{ry0}{ry0d}
      \xo{rx0}{rx0ry0d}
      \xo{ry0}{rx0ry0d}
    \end{tikzinline}}
\end{gather*}
In the diagram, the dotted lines indicate set inclusion into the index of the transformer-family.
As a quick correctness check, we can see that sequential composition is
associative in this case, since it does not matter whether we
associate to the left---with the intermediate step as in the diagram above,
eliding the write action---or to the right---with the intermediate step:
% \begin{gather*}
%   \PR{x}{r}\SEMI \PR{y}{s}
%   \\
%   \hbox{\begin{tikzinline}[node distance=2ex and 1.5em]
%       \xform{rx0d}{(0{=}r) \limplies \bForm}{}
%       \event{rx0}{\DR{x}{0}}{right=of rx0d}
%       \xform{rx0ry0d}{(0{=}r) \limplies(0{=}s) \limplies \bForm}{right=of rx0}
%       \event{ry0}{\DR{y}{0}}{right=of rx0ry0d}
%       \xform{ry0d}{(0{=}s) \limplies \bForm}{right=of ry0}
%       \xform{wz0i}{\bForm}{left=of rx0d}
%       \xo{rx0}{rx0d}
%       \xo{ry0}{ry0d}
%       \xo{rx0}{rx0ry0d}
%       \xo{ry0}{rx0ry0d}
%     \end{tikzinline}}
% \end{gather*}
\begin{gather*}
  \PR{y}{s}\SEMI \IF{s} \THEN \PW{z}{r{*}(s{-}1)} \FI
  \\
  \hbox{\begin{tikzinline}[node distance=1ex and 1.5em]
      \event{ry0}{\DR{y}{1}}{}
      \event{wz0}{(1{=}s) \limplies (s{\neq}0) \land (r{*}(s{-}1)){=}0 \bigmid \DW{z}{0}}{right=of ry0}
      \xform{ry0d}{(1{=}s) \limplies \bForm}{left=of ry0}
      \xform{wz0i}{\bForm}{left=of ry0d}
      \po{ry0}{wz0}
      \xo{ry0}{ry0d}
    \end{tikzinline}}
\end{gather*}
This is an instance of the general result that sequential composition forms a
monoid (\reflem{lem:monoid}).
